---
title: "4. Plotting stuff"
output: rmarkdown::html_vignette
date: "`r format(Sys.time(),  '%d.%m.%Y')`"
vignette: >
  %\VignetteIndexEntry{4. Plotting stuff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Plotting in `gen3sis2` is quite easy, as the package features several plotting functions to use in different contexts:

-   `plot_space_overview`
-   `plot_space`
-   `plot_species_presence`
-   `plot_abundance`
-   `plot_richness`

The core of plotting functions of `gen3sis2` is built with `ggplot2`. Among other things, this allows for plots to be easily customizable. In this vignette we will go trough an overwiew of `gen3sis2` plotting functions and how to customize your plot.

# Visualizing the space

## From you spaces.rds

If you inspect your spaces.rds file, you will see that it consists of an organized list of matrices and vectors, which aren't so good for visualization. To better visualize the spaces created, users can use the `plot_space_overview` function, that will plot one or more variable for defined timesteps.

Let's see an example:

```{r}
# Load some spaces.rds create with create_spaces_raster...
space <- readRDS(system.file("extdata/SouthAmerica/landscape/spaces.rds", package = "gen3sis2"))

# ... and plot it!
# "env_names" selects the variables of interest
# "breaks" selects the timesteps of interest
plot_space_overview(space, env_names = "temp", breaks = c(60, 50, 40, 30, 20, 10))
```

## From inside the simulation (called in the config)

The observer function is a great place to make plot calls, and `gen3sis2` features several functions to be called in this context. For example, it is possible to plot the space in the current timestep using the function `plot_space`.

As this vignette is for illustration purposes only, let's define a function that will alow us to simulation how your space looks like inside the simulation:

```{r}
simulate_space_format <- function(config,space,output_directory) {
  directories <- prepare_directories(config_file = config,
                                     input_directory = space,
                                     output_directory = output_directory)
  config <- create_input_config(config_file = config)
  config[["directories"]] <- directories
  val <- list("data" = list(),
              "vars" = list(),
              "config" = config)

  val$config <- gen3sis2:::complete_config(val$config)
  val$config$gen3sis$general$verbose <- 1
  val <- gen3sis2:::setup_inputs(val$config, val$data, val$vars)
  val <- gen3sis2:::setup_variables(val$config, val$data, val$vars)
  val <- gen3sis2:::setup_space(val$config, val$data, val$vars)
  return(val$data)
}

# 
config <- create_input_config(system.file("extdata/SouthAmerica/config/config_southamerica.R", package = "gen3sis2"))

data <- simulate_space_format(
  config = system.file("extdata/SouthAmerica/config/config_southamerica.R", package = "gen3sis2"),
  space = system.file("extdata/SouthAmerica/landscape", package = "gen3sis2"),
  output_directory = tempdir()
)

names(data)
```

[!] IMPORTANT: This is done just for this vignette, to facilitate the example.

Now, we can call the plot function in a very easy way:

```{r}
plot_space(data$space)
```

Again, this function is meant to be called *inside* the simulation and configured in the config file. `plot_space_overview`, however, is meant to be called *outside* the simulation, using the spaces.rds. 

# Visualizing species

## Presence

The `plot_species_presence` is an useful function that users can use to plot the species' presence across the space. To

```{r}
# Simulate species in random places
set.seed(13)
random_places <- sample(row.names(data$space$coordinates), 12)
sp <- gen3sis2::create_species(random_places, config)

# the same space we simulated before
plot_species_presence(sp, data$space)
```

## Abundance

```{r}
# Simulate species in random places...
sp <- gen3sis2::create_species(random_places, config)

# ... and with random abundance
sp$abundance[] <- sample(1:5, 12, replace = TRUE)

# the same space we simulated before
plot_species_abundance(sp, data$space)
```

The scale of the plot is adaptative. If the abundance vector of the species have few unique values, it creates a discrete scale. However, if the vector have a lot of unique values, it creates a continuous scale:

```{r}
# Simulate species in random places...
sp <- gen3sis2::create_species(random_places, config)

# ... and with random abundance
sp$abundance[] <- sample(1:200, 12, replace = TRUE)

# the same space we simulated before
plot_species_abundance(sp, data$space)
```

## Richness

Species richness across the space is a very useful information for users. For that reason, `gen3sis2` features the `plot_richness` function, which can be used within the config to plot species richness. Different of the other functions, this one will recieve a species list, not a single species. But, besides of that, it works in the very same way:

```{r}
# Simulating a species list
## inside the simulation this will be called "all_species"
species_list <- list(
  gen3sis2::create_species(random_places[1:10], config),
  gen3sis2::create_species(random_places[5:10], config),
  gen3sis2::create_species(random_places[8:12], config)
)

plot_richness(species_list, data$space)
```

# Plot aesthetics customization

Many aspects of the plot can be customized, ranging from simple to complex custom settings. This is possible due to the plotting core being built with `ggplot2`. Let's see now some possibilities.

## Customizing color

The simplest way to customize plots its defining colors, using the `col` parameter. Each context have its specifications, so the user needs to read functions' documentation to be aware of the proper usage. However, let's go over one simple example:

```{r}
# Simulate species in random places
set.seed(13)
random_places <- sample(row.names(data$space$coordinates), 12)
sp <- gen3sis2::create_species(random_places, config)

# It is possible to set any color, using names or hex codes
plot_species_presence(sp, data$space, col = c("lightgray","#8B0000"))
```

## Applying `ggplot2` syntax

As the plotting core of `gen3sis2` is the `ggplot2` package, plotting functions return ggplot2 objects. As a result, its natively possible to apply `ggplot2` syntax `gen3sis2` plots:

```{r}
# storing plot in an object 
p <- plot_species_presence(sp, data$space, col = c("black","#1B8200"))

# applying ggplot2 syntax
p +
  ggplot2::theme_dark() +
  ggplot2::labs(
    title = "A very informative map"
  )
```

## Customizing `gen3sis2` default aesthetics

Default plot aesthetics in `gen3sis2` are stored as the functions `raster_plot_aesthetics`, for raster type spaces, and `sf_plot_aesthetics`, for h3 and points type spaces. This allow users to customize the aesthetics from all plots in session, by masking the functions in the environment.

First, let's see how the `raster_plot_aesthetics` look like:

```{r}
raster_plot_aesthetics
```

Now, let's mask it, defining our own aesthetics:

```{r}
raster_plot_aesthetics <- function(space, col, x_breaks, y_breaks, title) {
  list(
    ggplot2::scale_x_continuous(
      expand = c(0, 0),
      breaks = x_breaks,
      labels = function(x) round(x, 0) # changed x axis labels...
    ),
    ggplot2::scale_y_continuous(
      expand = c(0, 0),
      breaks = y_breaks,
      labels = function(y) round(y, 0) # and y labels...
    ),
    ggplot2::theme_dark(), # and the theme...
    ggplot2::theme(
      plot.title = ggplot2::element_text(hjust = 0.5),
      panel.grid.major = ggplot2::element_line(color = alpha("white", 0.5), linewidth = 0.2), # and background lines...
      panel.grid.minor = ggplot2::element_line(color = alpha("white", 0.4), linewidth = 0.1),
      axis.text = ggplot2::element_text()
    ),
    ggplot2::labs(title = "Look, mom. It's custom!") # and title.
  )
}
```

Now the plots should have this new aesthetics:

```{r}
plot_species_presence(sp, data$space)
```

