---
title: "Implementing ecological and sympatric speciation"
author: "Bouwe Reijenga"
date: "2026-05-21"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A working definition for both ecological and sympatric speciation

*Sympatric speciation*: Speciation in which reproductive isolation evolves 
without geographic isolation; i.e., lineages split while occupying the same 
geographic area or while geographic barriers do not prevent gene flow. 
Bolnick & Fitzpatrick (2007) *AREES*, Coyne (2007) *Curr Biol*


*Ecological speciation*: a speciation process in which divergent natural selection
drives the evolution of reproductive incompatibility (i.e. isolation) between
taxa. Nosil *et al.* (2008) *TREE*

## Motivation

Currently the implementation of speciation in gen3sis is limited to allopatric 
speciation of geographically isolated clusters. The divergence factor functions 
are flexible enough to allow for factors such as the environment and species' 
traits to influence cluster divergence and speciation. However, speciation can't
happen within a cluster. This rules out the occurrence of sympatric speciation 
 and fine-scale ecological speciation modes.

## Limits on the current implementation

While highly flexible, the past and current version of gen3sis does not allow 
for within-cell ecological or sympatric speciation. It assumes that the 
units of speciation are (1) population-cell clusters, and (2) gene flow 
decreases divergence within population-cell clusters.


(i) Population-cell clusters, that operate as the unit of speciation, are 
determined by distance-based clustering using *Tdbscan_variable()*. The clusters 
are based on distances between cells and additionally the dispersal ability of
the species. This limit the frequency of single-cell speciation, and makes 
within-cell speciation not easily implementable by default.


```{r, eval = FALSE}
# part of the loop_speciation function (L55-L65). This part determines the 
# population-cell clusters for one species.
loop_speciation <- function(){
  
  # ...
  
  if ( length(species_presence)==1 ){#check if only one cell is occupied
    clu_geo_spi_ti <- 1
  }else{
    distances <- config$gen3sis$dispersal$get_dispersal_values(
      length(species_presence), species, data$space, config)

    permutation <- sample(1:length(species_presence), length(species_presence))
    clu_geo_spi_ti <- Tdbscan_variable(
      data$distance_matrix[species_presence[permutation],
                           species_presence[permutation],
                           drop=FALSE], distances, 1)
    clu_geo_spi_ti <- clu_geo_spi_ti[order(permutation)]
  }
  
  # ...
}
```

(i) Specifically, clusters become new species if they meet an a priori set
threshold:

```{r, eval = FALSE}

clu_gen_spi_ti_c <- Tdbscan(gen_dist_spi$compressed_matrix, 
                            config$gen3sis$speciation$divergence_threshold, 1)

```

(ii) Divergence between populations within the same cluster is decreased 
by "dfactor", whereas divergence increases between clusters based on the 
ifactor. The ifactor is derived from the user specified *get_divergence_factor()*
with a default value of 1, and can result in creative implementations such as 
divergence between clusters based on environmental factors or species' 
cluster-level traits. However, it does mean that populations within a cluster
by default do not diverge from one another. Even if we were to model frequency-
dependent competiton and/or divergent selection, a population can't split off.

-- Note to self: I don't understand why dfactor
is compensated for ifactor when dealing with a uniform ifactor, but not when
ifactor is supplied as a matrix?

```{r, eval = FALSE}
#' Updates a given divergence matrix
#'
#' @param gen_dist_spi a divergence matrix
#' @param clu_geo_spi_ti a cluster index
#' @param ifactor the divergence factor by which the clusters distances increase
#'
#' @return an updated divergence matrix
#' @noRd
update_divergence <- function(divergence, cluster_indices, ifactor) {
  # update genetic distance
  clusters <- unique(cluster_indices)
  if( length(ifactor) == 1 ) {
    # scalar ifactor
    divergence <- divergence + ifactor
    dfactor <- 1+ifactor
  } else {
    # matrix ifactor
    divergence <- divergence + ifactor[cluster_indices, cluster_indices]
    dfactor <- 1
  }
  for ( i in clusters ){
    # in case they belong to same clusters, subtract -2 (for the default case), 
    # so that the final difference is -1 given previous addition!
    divergence[cluster_indices == i, cluster_indices == i] <-
      divergence[cluster_indices == i, cluster_indices== i] - dfactor
  }
  # setting -1 to zero. Genetic differences can not be negative
  divergence[divergence < 0] <- 0
  ## end updating genetic distance ##
  return(divergence)
}
```

## Possible solutions

There are multiple solutions to the incorporation of ecological and sympatric 
speciation. Ideally we would like to make the framework as flexible as possible,
so that others might implement their own versions. The simplest incorporation
would be to let a given number of individuals in a cell become a new species. 
For this to be at all possible, the *loop_speciation()* would need to be updated 
by either adjusting *update_divergence()*, and/or incorporating a new function or 
loop that allows for within-cell speciation. e.g.,


```{r, eval = FALSE}
  within_population_speciation <- function(species, space, config, ...){
    # the function could return a new species but also may have to modify 
    # the ancestral population/species
  }
```

gen3sis2 already offers great flexibility with the storage of traits. These 
traits store information on a population basis. For instance, consider that not
only the mean of a ecologically-relevant trait is stored but also for example:

 - z_mean         = mean phenotype in the cell
 - z_within_sd    = within-mode standard deviation
 - z_mode_delta   = separation between two latent ecological modes
 - z_mode_p       = frequency of mode 1
 - z_bimodality   = derived index of how clearly two modes exist
 - z_internal_RI  = accumulated within-cell reproductive isolation
 - assortment     = strength of assortative mating / habitat fidelity

*z_internal_RI* could then be used to either probabilistically determine speciation
or by some a priori set threshold, being influenced by the trait distribution.
High bimodality of the trait may then result in high within-cell RI. upon 
speciation, so as output of *within_population_speciation()*, we might then see
a split between the species:

###### Before
- species A in cell i
  - abundance = 100
  - z_mode_p = 0.55
  - mode_1 = 0.2
  - mode_2 = 0.8

###### After:
- species A in cell i
  - abundance = 55
  - z_mean = 0.2
  - z_mode_delta = 0
- species B in cell i
  - abundance = 45
  - z_mean = 0.8
  - z_mode_delta = 0

Such an implementation would be preferential over, say, modelling multiple demes
per cell, as that would require a more extensive rework and may conflict with the
philosophy behind gen3sis. Alternative rules, perhaps making 
*within_population_speciation()* user specified (?) could follow:

- z_internal_RI > RI_threshold 
  - (where *z_internal_RI* may also be a function of 
    homogenisation in *update_divergence()*)
- z_bimodality > bimodality_threshold 
  - (strictly ecological trait based)
- z_mode_delta / z_within_sd > separation_threshold 
  - (are the distributions separated enough/non-overlapping?)
- min(z_mode_p, 1 - z_mode_p) > minimum_mode_frequency 
  - (minimum frequency of the two peaks)
- abundance * min(z_mode_p, 1 - z_mode_p) > minimum_mode_abundance 
  - (or abundance based)

*apply_ecology()* and *apply_evolution()* already offer staggering flexibility, 
as *apply_ecology* is supplied with abundance, traits, environment, and the 
config and *apply_trait_evolution()* with species, cluster_indices, space, and 
the config. *apply_ecology()* updates abundances and could do this for distinct 
modes for the trait distribution of a population. However, It is currently 
hard coded that *apply_ecology()* only returns abundances in *loop_ecology()*: 

```{r, eval = FALSE}
# L85 in loop_ecology()
NEW_abd <- config$gen3sis$ecology$apply_ecology(abundance, traits, 
                                                local_environment, config)
```

###### What I think therefore needs to be added or discussed:
- Speciation of a single species in a single cell into two
- Partitioning of abundance among the two derived lineages
- Not sure how gen3sis handles will handle such a phylogenetic split
  - i.e. which population remains with the rest of the range?
- create overlapping ranges in the same cell
- mode-specific dispersal? How does the trait distribution affect dispersers?
- *update_divergence()*, *loop_speciation()*, *loop_ecology()* and perhaps more

###### We would probably not be modelling:
 - Demes
 - Allele frequencies
 - Linkage disequilibrium
 - Hybridisation
 - Recombination
 - Mate choice

The implementation would thus not explicitly consider pop gen, and remain 
largely phenomenological. However, perhaps the Ne implementation can have 
a synergistic relationship.


