---
title: "Implementing ecological and sympatric speciation"
author: "Bouwe Reijenga"
date: "2026-05-21"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A working definition for both ecological and sympatric speciation

*Sympatric speciation*: "Sympatric speciation involves the splitting of an 
ancestral species into two or more reproductively isolated groups without 
geographical isolation of those groups. The key aspect of sympatric speciation 
is that it occurs when incipient species are in physical contact with each 
other, potentially able to interbreed and exchange genes." 
Coyne (2007) *Current Biology*


*Ecological speciation*: a speciation process in which divergent natural selection
drives the evolution of reproductive incompatibility (i.e. isolation) between
taxa. Nosil *et al.* (2008) *Trends in Ecology and Evolution*

## Motivation

Speciation is often classified based on the spatial context under which 
reproductive isolation evolves and the mechanism that drives the build up of 
reproductive isolation. The spatial context is mainly classified in four ways:
Under allopatric speciation, populations diverge after geographical isolation 
prevents or reduces gene flow; under peripatric speciation, isolated peripheral 
populations diverge; under parapatric speciation neighbouring populations diverge 
despite some spatial contact and gene flow; and under sympatric speciation, 
reproductive isolation evolves without geographic separation. Out of these 
scenarios, allopatric speciation is often seen as the most common and has been 
used as a baseline for the development of theoretical models such as gen3sis.

Spatial modes are distinct from the mechanisms generating reproductive 
isolation. Reproductive isolation may arise through ecologically based divergent selection 
(ecological speciation) or through non-ecological processes such as the fixation 
of different incompatible mutations under similar selection pressures 
(mutation-order speciation). The build up of reproductive isolation through ecological 
factors may happen while sympatric, after an initial phase of allopatry upon 
secondary contact between incipient species, fully in allopatry, or a mix of the above. 
Although genome sequences have revealed support for both sympatric and ecological 
speciation, ecological speciation is a less controversial route to speciation.

The specific spatial context and mechanism matters for emerging 
spatial and trait-based biodiversity patterns. For instance, sympatric speciation 
results in species that immediately show geographic range overlap and have 
diverged in important ecological traits such as their preferred food source. 
In contrast, allopatric modes of speciation will result in geographically 
isolated, largely ecological equivalent sister taxa. The build-up of diversity on a 
macroecological scale through time and space could therefore differ, but precise 
consequences are unclear without deeper theoretical study.

Currently, the implementation of speciation in gen3sis is limited to allopatric 
speciation, where geographically isolated clusters are the units that accumulate 
divergence. These clusters may differentiate based on environmental conditions 
and species traits, so ecological speciation can be modelled at the landscape/cluster 
scale. However, because divergence only accumulates between clusters and not within, 
the current implementation does not allow for ecological speciation at a finer 
spatial scale such as within a cluster, speciation at periphery of a species range, 
speciation under gene flow within a cluster (parapatry), or sympatric speciation 
within a single cell. 

```{r speciation_modes_plot, fig.width=7, fig.height=8, echo=FALSE}
# Base R schematic of speciation modes with touching hexagons

# ---------- helper functions ----------

hex_coords <- function(x, y, r = 0.5) {
  # pointy-top hexagon
  angles <- pi / 6 + seq(0, 2 * pi, length.out = 7)
  cbind(x + r * cos(angles), y + r * sin(angles))
}

draw_hex <- function(x, y, r = 0.5, border = "grey50", col = "grey95") {
  h <- hex_coords(x, y, r)
  polygon(h[, 1], h[, 2], border = border, col = col, lwd = 1.2)
}

individual_offsets <- function(scale = 0.18) {
  data.frame(
    dx = scale * c(-0.7,  0.0,  0.7, -0.35, 0.35, 0.0),
    dy = scale * c( 0.45, 0.65, 0.45, -0.35, -0.35, -0.7)
  )
}

draw_individuals <- function(x, y, state = rep("ancestral", 6)) {
  offs <- individual_offsets()

  cols <- ifelse(state == "derived", "red3", "black")
  pchs <- ifelse(state == "derived", 17, 16)

  points(
    x + offs$dx,
    y + offs$dy,
    pch = pchs,
    col = cols,
    cex = 1.1
  )
}

draw_panel <- function(title, cell_states, cells, hex_r) {
  plot(
    NA,
    xlim = c(min(cells$x) - 1.0, max(cells$x) + 1.2),
    ylim = c(min(cells$y) - 1.1, max(cells$y) + 1.1),
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    main = title
  )

  for (i in seq_len(nrow(cells))) {
    draw_hex(cells$x[i], cells$y[i], r = hex_r)
    draw_individuals(cells$x[i], cells$y[i], cell_states[[i]])
  }

  text(mean(cells$x[1:3]), min(cells$y) - 0.85, "Cluster A", cex = 0.8)
  text(mean(cells$x[4:5]), min(cells$y) - 0.85, "Cluster B", cex = 0.8)
}

# ---------- honeycomb geometry ----------

hex_r <- 0.5

# For pointy-top hexagons, adjacent centers that share an edge are:
dx <- sqrt(3) * hex_r
dy <- 1.5 * hex_r

# Three-cell cluster: all cells touch by edges
left_cluster <- data.frame(
  cell = paste0("L", 1:3),
  x = c(0, dx, dx / 2),
  y = c(0, 0, dy)
)

# Two-cell cluster: two touching cells
shift_x <- 5.0
right_cluster <- data.frame(
  cell = paste0("R", 1:2),
  x = shift_x + c(0, dx),
  y = c(0, 0)
)

cells <- rbind(left_cluster, right_cluster)

# ---------- define states ----------

ancestral <- rep("ancestral", 6)
derived   <- rep("derived", 6)
mixed     <- c(rep("derived", 3), rep("ancestral", 3))

initial_state <- list(
  ancestral, ancestral, ancestral,
  ancestral, ancestral
)

allopatric_state <- list(
  derived, derived, derived,
  ancestral, ancestral
)

within_cluster_state <- list(
  derived, derived, ancestral,
  ancestral, ancestral
)

within_cell_state <- list(
  mixed, ancestral, ancestral,
  ancestral, ancestral
)

# ---------- plot ----------

old_par <- par(no.readonly = TRUE)

par(
  mfrow = c(4, 1),
  mar = c(1.2, 1, 2.2, 1),
  oma = c(0, 0, 0, 0)
)

draw_panel(
  "(i) Initial state: same species in all cells",
  initial_state,
  cells, hex_r
)

draw_panel(
  "(ii) Allopatric speciation: isolated cluster changes species",
  allopatric_state,
  cells, hex_r
)

draw_panel(
  "(iii) Within-cluster speciation: part of one cluster changes species",
  within_cluster_state,
  cells, hex_r
)

draw_panel(
  "(iv) Within-cell speciation: part of one cell changes species",
  within_cell_state,
  cells, hex_r
)

par(old_par)
```

**Figure 1** Spatial modes of speciation. The figure shows the different modes of 
speciation in a spatial cell-based context. (i) The initial state prior to 
speciation. The two clusters are isolated and therefore build up divergence over 
time. (ii) In the current implementation of gen3sis, these two clusters may become 
different species. Because of the current restrictions (iii) within-cluster speciation, 
e.g. parapatry, is not possible, nor is (iv) within-cell sympatric speciation.

## Limits on the current implementation

While highly flexible, the past and current version of gen3sis assumes that (1) the 
units of speciation are population-cell clusters, that (2) gene flow 
decreases divergence within population-cell clusters purely based on being part 
of the same geographical cluster (but see (ii) below), and (3) that these clusters are determined 
based on dispersal ability (e.g. not affected by large trait differences that 
may affect hybrid fitness, important in ecological speciation).


(i) Population-cell clusters, that operate as the unit of speciation, are 
determined by distance-based clustering using *Tdbscan_variable()*. The clusters 
are based on distances between cells and the dispersal ability of
the populations (potentially informed by traits and the environment).
This limits the occurrence of peripheral speciation unless the populations are 
separated in space and beyond dispersal kernels. Additionally, within-cell 
speciation is currently not possible.

Note that *config\$gen3sis\$dispersal\$get_dispersal_values()* is user specified. 
While the dispersal kernels of individual populations (and cells) can be made 
dependent on the trait values of the population, i.e. populations can be made to 
disperse further based on their trait value, dispersal and by extension gene flow 
cannot be reduced specifically between populations that differ strongly in trait value. 
As a result, ecological reproductive barriers that depend on pairwise
trait divergence among populations cannot affect the formation of reproductive
clusters under the current clustering implementation.

```{r, eval = FALSE}
# part of the loop_speciation function (L55-L65). This part determines the 
# population-cell clusters for one species.
loop_speciation <- function(){
  
  # ...
  
  if ( length(species_presence)==1 ){#check if only one cell is occupied
    clu_geo_spi_ti <- 1
  }else{
    # obtain the dispersal kernels that are handed to the cluster-function
    distances <- config$gen3sis$dispersal$get_dispersal_values(
      length(species_presence), species, data$space, config)

    permutation <- sample(1:length(species_presence), length(species_presence))
    # cluster based on distances between cells and kernels
    clu_geo_spi_ti <- Tdbscan_variable(
      data$distance_matrix[species_presence[permutation],
                           species_presence[permutation],
                           drop=FALSE], distances, 1)
    clu_geo_spi_ti <- clu_geo_spi_ti[order(permutation)]
  }
  
  # ...
}
```

Clusters become new species if they meet an *a priori* set
threshold *config\$gen3sis\$speciation\$divergence_threshold*:

```{r, eval = FALSE}

clu_gen_spi_ti_c <- Tdbscan(gen_dist_spi$compressed_matrix, 
                            config$gen3sis$speciation$divergence_threshold, 1)

```

(ii) Divergence between populations within the same cluster is decreased 
by *dfactor*, whereas divergence increases between clusters based on the 
*ifactor*. The *ifactor* is derived from the user specified *get_divergence_factor()*
with a default value of 1, and can result in creative implementations such as 
divergence between clusters based on environmental factors or species' 
cluster-level traits. However, it does mean that populations within a cluster
by default do not diverge from one another (although there may be some flexibility 
if an ifactor-matrix is used; see comment below). Even if we were to model 
frequency-dependent competiton, divergent selection, assortative mating, etc., a 
within-cluster speciation event cannot occur.

Note that if *ifactor* is a matrix, and that divergence happens distinctly between 
clusters, e.g. based on traits, the diagonal can be used to compensate for convergence 
of cells within a cluster. Specifically, the user is responsible for keeping the 
diagonal at 0, or <1 for slower convergence, or <0 for faster convergence. While not 
directly available now, this can be used to track the possibility of within-cluster 
speciation in the future, because convergence could be made dependent on the trait 
or environmental differences within a cluster.

```{r, eval = FALSE}
#' Updates a given divergence matrix
#'
#' @param gen_dist_spi a divergence matrix
#' @param clu_geo_spi_ti a cluster index
#' @param ifactor the divergence factor by which the clusters distances increase
#'
#' @return an updated divergence matrix
#' @noRd
update_divergence <- function(divergence, cluster_indices, ifactor) {
  # update genetic distance
  clusters <- unique(cluster_indices)
  if( length(ifactor) == 1 ) {
    # scalar ifactor
    divergence <- divergence + ifactor
    dfactor <- 1+ifactor
  } else {
    # matrix ifactor
    divergence <- divergence + ifactor[cluster_indices, cluster_indices]
    dfactor <- 1
  }
  for ( i in clusters ){
    # in case they belong to same clusters, subtract -2 (for the default case), 
    # so that the final difference is -1 given previous addition!
    divergence[cluster_indices == i, cluster_indices == i] <-
      divergence[cluster_indices == i, cluster_indices== i] - dfactor
  }
  # setting -1 to zero. Genetic differences can not be negative
  divergence[divergence < 0] <- 0
  ## end updating genetic distance ##
  return(divergence)
}
```

## Possible solutions

There are multiple solutions to the incorporation of ecological, peripetric, 
parapatric, and sympatric speciation. Ideally we would like to make the framework 
as flexible as possible, so that others might implement their own versions. There 
are three ways of speciation that could be incorporated (Figure 1): (1) the main 
between-cluster speciation as currently incorporated, (2) The within-cluster 
speciation that splits of multiple cells at once (i.e. speciation with gene flow), 
and (3) within-cell speciation that could mimic point-speciation or sympatric 
speciation (i.e. the ancestor is sympatric with the new species). 
Out of these options I explore an ecologically-motivated example of the latter 
below, as it would likely require a new function (e.g. (2) could be an extension 
of the current cluster-based speciation although focusing on trait-based 
clusters instead). 

The simplest incorporation of sympatric speciation would be to let a given number 
of individuals in a cell become a new species. For this to be at all possible, 
the *loop_speciation()* would need to be updated by either adjusting *update_divergence()*, 
and/or incorporating a new function or loop that allows for within-cell speciation. 
We think that the best solution, however, is to add a separate function that exists 
in parallel to the current allopatric speciation loop. e.g.,


```{r, eval = FALSE}
  within_population_speciation <- function(species, space, config, ...){
    # the function could return a new species but also may have to modify 
    # the ancestral population/species
  }
```

gen3sis2 already offers great flexibility with the storage of traits. These 
traits store information on a population basis. For instance, consider that not
only the mean of a ecologically-relevant trait is stored but also for example:

 - z_mean         = mean phenotype in the cell
 - z_within_sd    = within-mode standard deviation
 - z_mode_delta   = separation between two latent ecological modes
 - z_mode_p       = frequency of mode 1
 - z_bimodality   = derived index of how clearly two modes exist
 - z_internal_RI  = accumulated within-cell reproductive isolation
 - assortment     = strength of assortative mating / habitat fidelity

*z_internal_RI* could then be used to either probabilistically determine speciation
or by some a priori set threshold, being influenced by the trait distribution.
High bimodality of the trait may then result in high within-cell RI. upon 
speciation, so as output of *within_population_speciation()*, we might then see
a split between the species:

###### Before
- species A in cell i
  - abundance = 100
  - z_mode_p = 0.55
  - mode_1 = 0.2
  - mode_2 = 0.8

###### After:
- species A in cell i
  - abundance = 55
  - z_mean = 0.2
  - z_mode_delta = 0
- species B in cell i
  - abundance = 45
  - z_mean = 0.8
  - z_mode_delta = 0

Such an implementation would be preferential over, say, modelling multiple demes
(or habitats) per cell, as that would require a more extensive rework and may conflict with the
philosophy behind gen3sis. Alternative rules, perhaps making 
*within_population_speciation()* user specified (?) could follow:

- z_internal_RI > RI_threshold 
  - (where *z_internal_RI* may also be a function of 
    homogenisation in *update_divergence()*)
- z_bimodality > bimodality_threshold 
  - (strictly ecological trait based)
- z_mode_delta / z_within_sd > separation_threshold 
  - (are the distributions separated enough/non-overlapping?)
- min(z_mode_p, 1 - z_mode_p) > minimum_mode_frequency 
  - (minimum frequency of the two peaks)
- abundance * min(z_mode_p, 1 - z_mode_p) > minimum_mode_abundance 
  - (or abundance based)

*apply_ecology()* and *apply_trait_evolution()* already offer flexibility to calculate 
these metrics, as *apply_ecology* is supplied with abundance, traits, environment, 
and the config and *apply_trait_evolution()* with species, cluster_indices, space, and 
the config. *apply_ecology()* updates abundances and could do this for distinct 
modes for the trait distribution of a population. However, It is currently 
hard coded, and expected from other functions, that *apply_ecology()* only returns 
abundances in *loop_ecology()*. This is problematic if we are to model, e.g., 
divergent selection that causes changes in abundances and trait values at the same 
instance.

###### What I think therefore needs to be added:
- Speciation of a single species in a single cell into two
- Partitioning of abundance among the two derived lineages
- Clarfiy how gen3sis will handle such phylogenetic splits
  - i.e. which population remains with the rest of the range?
  - How do speciation modes and the loops interact: Do we get allopatric speciation 
  immediately following sympatric speciation?
- Consider overlapping ranges in the same cell
- mode-specific dispersal? How does the trait distribution affect dispersers?
- *update_divergence()*, *loop_speciation()*, *loop_ecology()* and perhaps more 
functions require to be updated.

###### We may not be modelling:
 - Demes
 - Allele frequencies
 - Linkage disequilibrium
 - Hybridisation
 - Recombination
 - Mate choice

The implementation would thus not explicitly consider population genetics, and remain 
largely phenomenological. However, perhaps the Ne implementation can have 
a synergistic relationship.


